---
title: "Lab9"
output:
  ioslides_presentation:
    widescreen: yes
    size: smaller
    font_adjustment: -1
    toc: 1
  beamer_presentation: default
date: "2022-11-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE,fig.align = 'center')
```

```{r, eval=T, echo=F}
suppressPackageStartupMessages(library(dplyr))
```

## Tasks

-   Loops: `for()`, `while()`
-   Loop-related: `apply`, `lapply`,`sapply`, `replicate`
-   Sampling in R: `sample`, `sample_n`, `rep_sample_n`
-   Use one or more of these tools to solve [Exercise 3 (Lab 6)](https://moodle.brynmawr.edu/pluginfile.php/267053/mod_folder/content/0/Lab_Guide6.pdf)
-   More on conditional statements
-   [Lab 9 Exercise 1](lab9_ex1.Rmd) and [Lab 9 Exercise 2](lab9_ex2.Rmd)
-   R Quiz 9: two chapters from Intermediate R on DataCamp ("Conditionals and Control Flow" and "Loops")

## Conditional statements

The main conditional statements in R are:

-   `if`: we saw this in Lab 8
-   `if`; `else`: we saw this in Lab 8
-   `else if`: new!

## Conditional statements: `else if`

Basic syntax

```{r, eval=F}
if(condition){ #if condition is true
  statement #do something
}else if(condition 2){
  statement 2 #do this other thing
}else{ #if neither statement nor statement 2 are true
  statement 3 # do this other other thing
}
```

## Conditional statements: `else if`

```{r}
lunch<-"Dog food"
if(lunch=="Pizza"){ #if first statatement
  print("Yay!")
}else if(lunch=="Rice & beans"){
  print("Sounds good!")
}else{
  print("Meh")
}
```

## Conditional statements: `else if`

```{r}
#you can use negatives too
lunch<-"Dog food"
if(lunch!="Pizza" & lunch!="Rice & beans"){ #if first statatement
  print("Meh")
}else if(lunch=="Rice & beans"){
 print("Sounds good!")
}else{
  print("Yay!")
}
```

## Loops in R

Three kinds of loops:

* `for` loop

* `while` loop 

* `repeat` loop [we won't discuss]





## Loops in R: `for` loops

Basic syntax

```{r, eval=F}
for (value in sequence){
  
  statement #DO SOMETHING
}

```

## Loops in R: `for` loops

```{r}
# iterate over elements of a sequence
for (i in 1: 4){
  
    print(i * 2)
  
}
```

## Loops in R: `for` loops

```{r}
#iterate over elements of a vector
for (i in c(-8^2, 2*4, 239, 29321)){

    print(i)
  
}
```

## Loops in R: `for` loops

```{r}
#create vector outside of loop and then use it to iterate
x <- c(-8^2, 2*4, 239, 29321)
for (i in x){
  
    print(i)
  
}
```

## Loops in R: `while` loops

Basic syntax

```{r, eval=F}
while ( condition ) {
  
  statement #do something
}

```


## Loops in R: `while` loops

```{r, eval=T}
y<-0
while (y <= 5){ #logical
    # statements
    print(y) # if y <=5, print it
    y = y + 1 #if y <=5 add 1 to it
}
y
```

## Loops in R: `while` loops

```{r}
# whose factorial will be calculated
n <- 5 
factorial <- 1 # assigning the factorial variable
i <- 1 # and iteration variable to 1
# using while loop
while (i <= n){
    # multiplying the factorial variable
    # with the iteration variable
    factorial = factorial * i
    # incrementing the iteration variable
    i = i + 1
}
print(factorial) # displaying the factorial
```

## Looping in the command line

`lapply()`: Loop over a `list` and evaluate a function on each element

`sapply()`: Same as `lapply` but try to simplify the result

`apply()`: Apply a function over the margins of an array

There are others, like `tapply()`, `mapply()`, but we won't discuss them.

## Looping in the command line:`lapply()`

-   it loops over a list, iterating over each element in that list
-   it applies a function to each element of the list (a function that you specify)
-   and returns a list (the `l` is for "list").

## Looping in the command line:`lapply()`

Basic syntax for `lapply` and `sapply`

```{r, eval=F}
lapply(X,FUN)
lapply(Data, Function to apply to data)
```

## Looping in the command line:`lapply()`

```{r}
#rnorm samples numbers randomly from a normal distribution
set.seed(123)
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
#str(x)
lapply(x, mean)
```

## Looping in the command line:`sapply()`

```{r}
#rnorm samples numbers randomly from a normal distribution
set.seed(123)
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
#str(x)
sapply(x, mean)
```

## Looping in the command line: `apply()`

Basic syntax

```{r, eval=F}

apply(X,MARGIN, FUN)
lapply(Data, Where the Function will be applied, Function to apply to data)
#Margin: 1 is rows
#Margin: 2 is cols
```

## Looping in the command line: `apply()`

```{r}
set.seed(123)
#make a matrix
mat<-matrix(rnorm(20,1), ncol=2)
apply(mat,1, mean) # 1 signifies rows
rowMeans(mat) #equivalent
```

## Looping in the command line: `apply()`

```{r}
set.seed(123)
#make a matrix
mat<-matrix(rnorm(20,1), ncol=2)
apply(mat,2, mean) # signifies columns
colMeans(mat) #equivalent
```

## The replicate function

```{r}
#replicate the value 3 exactly 10 times
replicate(n=10, 3)
#replicate the letter 'A' exactly 7 times
replicate(n=7, 'A')
```

## The replicate function

```{r}
set.seed(1)
#generate 3 values that follow normal distribution (replicate this 4 times)
replicate(n=4, rnorm(3, mean=0, sd=1))
```

## Practice!

[Lab9_ex2.Rmd](Lab9_ex2.Rmd)

## Sampling in R

`sample(x = , size = , replace = , prob = )`: Generate a sample of size size, from a vector `x`, with `(replace = TRUE)` or without `(replacement = FALSE)` replacement. By default the size is the length of `x`, sampling occurs without replacement and probabilities are equal. Change these defaults by specifying a value for the argument. For example, to have unequal sampling probabilities, include a vector of length `x`, in which the $i^th$ entry describes the relative probability of sampling the $i^th$ value in `x`.

`sample_n(tbl = , size = , replace = , weight = )`: Generate a sample of size `size`, from a `tibble tbl`, with `(replace = TRUE)` or without `(replacement = FALSE)` replacement. All arguments are the same as in `sample()` except `weight` replaces `prob`, and `tbl` replaces `x`. `sample_n()` is a function in the `dplyr` package, which is loaded with `tidyverse`.

## New functions for sampling in R!

These functions extend the functionality of dplyr::sample_n() and dplyr::slice_sample() by allowing for repeated sampling of data. This operation is especially helpful while creating sampling distributions

These are related to `dplyr::sample_n` and \`dplyr::slice_sample, respectively:

`rep_sample_n():` Generate `reps` samples of size `size`

`rep_sample_slice():` Generate `reps` samples of size `n`

Both return a tibble grouped by replicate. \`

## New functions for sampling in R!

```{r}
library(infer) # a tidymodels package
#sample 2 rows from the iris dataset without replacement. Repeat this 2 times
set.seed(123)
test1<-iris %>% 
  rep_sample_n(size = 3, replace = F, reps = 2) #uses size instead of n
test1
```

## New functions for sampling in R!

```{r}
library(infer) # a tidymodels package
#sample 2 rows from the iris dataset without replacement. Repeat this 2 times
set.seed(123)
test2<-iris %>% 
  rep_slice_sample(n = 3, replace = F, reps = 2) #uses n instead of size
test2
```

## New functions for sampling in R!

```{r}
test1==test2
```

## Back to the Exercise: Making a distribution of sample means (from lab 6)

Last week you learned how to take a sample (random rows) from a dataset and calculate statistics on it.

Specifically, you took two independent samples from the `human_genes` dataset and calculated some descriptive statistics.

## Exercise: Making a distribution of sample means

Step 1: Read in the `human_genes.csv` dataset and get summaries

```{r}
library(dplyr) #load the dplyr package
library(readr) #load the readr package
#read in the human genes dataset
human_genes<-readr::read_csv("input_files/human_genes.csv")
```

## Exercise: Making a distribution of sample means

Step 1: Read in the `human_genes.csv` dataset and get summaries

```{r}
glimpse(human_genes) #glimpse the dataset
#remove useless columns
human_genes<-human_genes %>% 
  select(name, size)
```

## Exercise: Making a distribution of sample means
Step 2: Calculate the following descriptive statistics for the dataset: `mean`, `median`,`standard deviation`, `mode`, `IQR`.

```{r}
#summarise the human genes dataset creating three columns
human_genes_summ <- human_genes %>%
  summarise(
  MeanLength = mean(size),
  MedianLength = median(size),
  SDLength = sd(size),
  IQR = IQR(size)
)
human_genes_summ #
```

## Exercise: Making a distribution of sample means
Step 3: take 10,000 random samples of length `n=100` from the `human_genes` dataset and calculate the same summaries as above.

Wow! Let's break this down:

-   sample 10 random samples of length `n=100`
-   calculate the same summaries as above for each sample
-   if everything works, switch to 10,000

## Exercise: Making a distribution of sample means
Step 3a: sample 10 random samples of length `n=100` (using `for`)

```{r}
library(dplyr)
x<-seq(1:10) #create a vector with numbers from 1 to 10000
# for each iteration,take a sample of size 100 without replacement from human_genes
samps<-list() #create empty list to put resamps in
for(i in x){ #for loop
  samps[[i]]<-human_genes %>% 
    sample_n(size=100, replace=F)
  print(i) #print each iteration 
}
```

## Exercise: Making a distribution of sample means

Step 3b: calculate the same summaries as above for each sample

```{r}
res<-list() #create empty list to put the summaries in
for(i in 1:length(samps)){ #for loop for each elements in the samps list
  res[[i]]<- samps[[i]] %>% 
   summarise(
    MeanLength = mean(size),
    MedianLength = median(size),
    SDLength = sd(size),
    IQR = IQR(size)
)
  print(i) #print each iteration 
}
res
```

## Exercise: Making a distribution of sample means

Can we do it all in one go?

```{r}
x<-seq(1:10) #create a vector with numbers from 1 to 10000
#create tibble with the columns produced by our summaries
#dummy tibble
resamps<-tibble(MeanLength=NA, MedianLength=NA, SDLength=NA, IQR=NA) 
for(i in x){ #for loop
  temp<-human_genes %>% #assign to temp object
    sample_n(size=100, replace=F) %>% #sample
    summarise( # summaries
    MeanLength = mean(size),
    MedianLength = median(size),
    SDLength = sd(size),
    IQR = IQR(size)
    )
  resamps<-bind_rows(resamps, temp) #add new row to tibble
  print(i) #print each iteration 
}
```

## Exercise: Making a distribution of sample means

```{r}
#the first row has NAs. we can easily get rid of it.

resamps<-na.omit(resamps) #remove lines with NA
resamps
```

## Exercise: Making a distribution of sample means

Step 3c: if everything works, switch to 10,000

```{r, eval=F}
nsamps=10000
resamps<-tibble(MeanLength=NA, MedianLength=NA, SDLength=NA, IQR=NA) 
for(i in 1:nsamps){ #for loop
  temp<-human_genes %>% #assign to temp object
    sample_n(size=100, replace=F) %>% #sample
    summarise( # summaries
    MeanLength = mean(size),
    MedianLength = median(size),
    SDLength = sd(size),
    IQR = IQR(size)
    )
  resamps<-bind_rows(resamps, temp) #add new row to tibble
  if(i %in%  seq(from=1, to=nsamps, by=500)){
  print(i) #print 1st,501st,1001st etc iterations
  }
}
```

## Exercise: Making a distribution of sample means

Another option: using `rep_slice_sample`

Make a function that calculates all the summaries

```{r}
library(infer)
#samps
nsamps<-10 #once this works, switch to 10,000
reps<-rep_sample_n(human_genes, size=100, replace=FALSE, reps=nsamps) %>% 
  group_by(replicate) %>% 
  summarise( # summaries
  MeanLength = mean(size),
  MedianLength = median(size),
  SDLength = sd(size),
  IQR = IQR(size)
) %>%
  ungroup
```

## Exercise: Making a distribution of sample means

Step 4: Plot the sampling distribution with ggplot

```{r, fig.height=2}
library(ggplot2)
#rerun previous code with nsamps=10000
ggplot(reps, aes(x=MeanLength)) + 
  geom_histogram() + 
  theme_minimal()

```

## Exercise: Making a distribution of sample means

Final step: Obtain the grand mean (mean of sample means) and the standard deviation of the sample means and: compare with the population mean; calculate SEM using the distribution and the formula and compare.

```{r}
#Try it!

```

## Exercises!


-   Binomial hypothesis testing: `lab9_ex1.Rmd`
-   Loops & Conditional Statements: Exercises in `lab9_ex2.Rmd`.
