---
title: "Lab9"
output:
  ioslides_presentation:
    widescreen: true
    size: smaller
    font_adjustment: -1
    toc: 1
date: "2022-11-02"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE,fig.align = 'center')
```

```{r, eval=T, echo=F}
suppressPackageStartupMessages(library(dplyr))
```

## Tasks

* Recap
* Conditional statements (cont.): else if/while
* Loops
* `apply`, `lapply`,`sapply`, `replicate`, `rep_sample_n` 
* Sampling in R: `sample`, `sample_n`, `rep_sample_n`
* Use one or more of these tools to solve Exercise 3 (Lab 7)
* Lab 9 Problems
* R Quiz 9 (will be posted tomorrow)

## Recap

### General functions:


`group_by()`: Conduct operations separately by values of a column (or columns).

`summarise()`: Reduces our data from the many observations for each variable to just the summaries we ask for. Summaries will be one row long if we have not group_by() anything, or the number of groups if we have.

`sum()`: Adding up all values in a vector.

`diff()`: Subtract sequential entries in a vector.

`sqrt()`: Find the square root of all entries in a vector.

`unique()`: Reduce a vector to only its unique values.

`pull()`: Extract a column from a tibble as a vector.

`round()`: Round values in a vector to the specified number of digits.



## Recap

### Summarizing location (center):


`n()`: The size of a sample.

`mean()`: The mean of a variable in our sample.

`median()`: The median of a variable in our sample.



### Summarizing width (spread):

`max()`: The largest value in a vector.

`min()`: The smallest value in a vector.

`range()`: Find the smallest and larges values in a vector. Combine with diff() to find the difference between the largest and smallest value.

`quantile()`: Find values in a give quantile. Use `diff(quantile(... , probs = c(0.25, 0.75)))` to find the interquartile range.

`IQR()`: Find the difference between the third and first quartile (aka the interquartile range).

`var()`: Find the sample variance of vector.

`sd()`: Find the sample standard deviation of vector.



## Conditional statements: else if

Last week we learned about `if` and `else`.

You can also


## Conditional statements

## Loops in R

Three kinds of loops:

*`for` loop
*`while` loop
*`repeat` loop


Resources [optional]:

* https://www.geeksforgeeks.org/loops-in-r-for-while-repeat/
* DataCamp: Intermediate R 


## Loops in R

```{r, eval=T, echo=F}
url<-'https://media.geeksforgeeks.org/wp-content/uploads/20200327192448/for_flowdiagram.png'
```


## Loops in R: `for` loops

Basic syntax

```{r, eval=F}
for (value in sequence){
  
  statement #DO SOMETHING
}

```


## Loops in R: `for` loops

![](`r url`)




## Loops in R: `for` loops

```{r}
# iterate over elements of a sequence
for (i in 1: 4){
  
    print(i * 2)
  
}
```


## Loops in R: `for` loops

```{r}
#iterate over elements of a vector
for (i in c(-8^2, 2*4, 239, 29321)){

    print(i)
}
```

## Loops in R: `for` loops

```{r}
#create vector outside of loop and then use it to iterate
x <- c(-8^2, 2*4, 239, 29321)

for (i in x){
  
    print(i)
}
```


## Loops in R: `while` loops

Basic syntax

```{r, eval=F}

while ( condition ) {
  
  statement #do something
}

```


## Loops in R: `while` loops

```{r, echo=F}
url2<-'https://media.geeksforgeeks.org/wp-content/uploads/20200327192542/while_flowdiagram.png'
```

![](`r url2`)


## Loops in R: `while` loops

```{r, eval=T}
y<-0

while (y <= 5){ #logical
    # statements
    print(y) # if y <=5, print it
    y = y + 1 #if y <=5 add 1 to it
}

y
```

## Loops in R: `while` loops

```{r}
# whose factorial will be calculated
n <- 5
# assigning the factorial variable
# and iteration variable to 1
factorial <- 1
i <- 1
# using while loop
while (i <= n){
 
    # multiplying the factorial variable
    # with the iteration variable
    factorial = factorial * i
 
    # incrementing the iteration variable
    i = i + 1
}
# displaying the factorial
print(factorial)
```



## Looping in the command line


`lapply()`: Loop over a `list` and evaluate a function on each element

`sapply()`: Same as `lapply` but try to simplify the result

`apply()`: Apply a function over the margins of an array

There are others, like `tapply()`, `mapply()`, but we won't discuss them.


## Looping in the command line:`lapply()`

* it loops over a list, iterating over each element in that list
* it applies a function to each element of the list (a function that you specify)
* and returns a list (the `l` is for “list”).

## Looping in the command line:`lapply()`

Basic syntax

```{r, eval=F}

lapply(data, function to apply to data)
```

## Looping in the command line:`lapply()`

```{r}
#rnorm samples numbers randomly from a normal distribution
set.seed(123)
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
str(x)
lapply(x, mean)
```

## Looping in the command line:`sapply()`

```{r}
#rnorm samples numbers randomly from a normal distribution
set.seed(123)
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
str(x)
sapply(x, mean)
```

## Looping in the command line: `apply()`

```{r}
set.seed(123)
#make a matrix
mat<-matrix(rnorm(20,1), ncol=2)

apply(mat,1, mean) # 1 signifies rows

apply(mat,2, mean) # signifies columns
```


## Sampling in R

`sample(x = , size = , replace = , prob = )`: Generate a sample of size size, from a vector `x`, with `(replace = TRUE)` or without `(replacement = FALSE)` replacement. By default the size is the length of `x`, sampling occurs without replacement and probabilities are equal. Change these defaults by specifying a value for the argument. For example, to have unequal sampling probabilities, include a vector of length `x`, in which the  $i^th$ entry describes the relative probability of sampling the  $i^th$ value in `x`. 

`sample_n(tbl = , size = , replace = , weight = )`: Generate a sample of size `size`, from a `tibble tbl`, with `(replace = TRUE)` or without `(replacement = FALSE)` replacement. All arguments are the same as in `sample()` except `weight` replaces `prob`, and `tbl` replaces `x`. `sample_n()` is a function in the `dplyr` package, which is loaded with `tidyverse`.

## Sampling in R: The replicate function:

```{r}
#replicate the value 3 exactly 10 times
replicate(n=10, 3)
#replicate the letter 'A' exactly 7 times
replicate(n=7, 'A')
```

## A new function for sampling in R!

`rep_sample_n():` Generate `reps` samples of size `siz`e, from a `tibble tb`l, with `(replace = TRUE)` or without `(replacement = FALSE)` replacement. All arguments are the same as in `sample()`, except `tbl` replaces `x`, and `reps` allows us to specify how many samples of size `size` we would like. `rep_sample_n()` returns a tibble grouped by replicate. `rep_sample_n()` is a function in the infer package, which must be installed and loaded in addition to tidyverse.


```{r}
#show example

```

## Back to the Exercise: Making a distribution of sample means (from lab 7)


Last week you learned how to take a sample (random rows) from a dataset and calculate statistics on it. 

Specifically, you took two independent samples from the `human_genes` dataset and calculated some descriptive statistics. 

## Exercise: Making a distribution of sample means (from lab 7)

Step 1: Read in the `human_genes.csv` dataset and get summaries

```{r}
library(dplyr) #load the dplyr package
library(readr) #load the readr package
human_genes<-readr::read_csv("input_files/human_genes.csv") #read in the human genes dataset
glimpse(human_genes) #glimpse the dataset
#remove useless columns
human_genes<-human_genes %>% 
  select(name, size)
```

## Exercise: Making a distribution of sample means (from lab 7)

Step 2: Calculate the following descriptive statistics for the dataset: `mean`, `median`,`standard deviation`, `mode`, `IQR`.

```{r}
#summarise the human genes dataset creating three columns
human_genes_summ <- human_genes %>%
  summarise(
  MeanLength = mean(size),
  MedianLength = median(size),
  SDLength = sd(size),
  IQR = IQR(size)
)
human_genes_summ #
```

## Exercise: Making a distribution of sample means (from lab 7)

Step 3: take 10,000 random samples of length `n=100` from the `human_genes` dataset and calculate the same summaries as above.


Wow! Let's break this down:

* sample 10 random samples of length `n=100`
* calculate the same summaries as above for each sample
* if everything works, switch to 10,000

## Exercise: Making a distribution of sample means (from lab 7)

Step 3a: sample 10 random samples of length `n=100` (using `for`)

```{r}
library(dplyr)
x<-seq(1:10) #create a vector with numbers from 1 to 10000
# for each iteration,take a sample of size 100 without replacement from human_genes
samps<-list() #create empty list to put resamps in
for(i in x){ #for loop
  
  samps[[i]]<-human_genes %>% 
    sample_n(size=100, replace=F)
  
  print(i) #print each iteration 
}
```

## Exercise: Making a distribution of sample means (from lab 7)

Step 3b: calculate the same summaries as above for each sample
```{r}
res<-list() #create empty list to put the summaries in
for(i in 1:length(samps)){ #for loop for each elements in the samps list
  
  res[[i]]<- samps[[i]] %>% 
   summarise(
    MeanLength = mean(size),
    MedianLength = median(size),
    SDLength = sd(size),
    IQR = IQR(size)
)
  
  print(i) #print each iteration 
}
res
```

## Exercise: Making a distribution of sample means (from lab 7)

Can we do it all in one go?

```{r}
x<-seq(1:10) #create a vector with numbers from 1 to 10000
#create tibble with the columns produced by our summaries
#dummy tibble

resamps<-tibble(MeanLength=NA, MedianLength=NA, SDLength=NA, IQR=NA) 
for(i in x){ #for loop
  
  temp<-human_genes %>% #assign to temp object
    sample_n(size=100, replace=F) %>% #sample
    summarise( # summaries
    MeanLength = mean(size),
    MedianLength = median(size),
    SDLength = sd(size),
    IQR = IQR(size)
    )

  resamps<-bind_rows(resamps, temp) #add new row to tibble
  print(i) #print each iteration 
}
```

## Exercise: Making a distribution of sample means (from lab 7)

```{r}
#the first row has NAs. we can easily get rid of it.

resamps<-na.omit(resamps)
resamps
```

## Exercise: Making a distribution of sample means (from lab 7)
Step 3c: if everything works, switch to 10,000

```{r}
nsamps=10000
resamps<-tibble(MeanLength=NA, MedianLength=NA, SDLength=NA, IQR=NA) 
for(i in 1:nsamps){ #for loop
  
  temp<-human_genes %>% #assign to temp object
    sample_n(size=100, replace=F) %>% #sample
    summarise( # summaries
    MeanLength = mean(size),
    MedianLength = median(size),
    SDLength = sd(size),
    IQR = IQR(size)
    )
  resamps<-bind_rows(resamps, temp) #add new row to tibble
  if(i %in%  seq(from=1, to=nsamps, by=100))
  print(i) #print each iteration 
}
```

## Exercise: Making a distribution of sample means (from lab 7)

Another option: using `rep_slice_sample`

Make a function that calculates all the summaries

```{r}
library(infer)
#samps
nsamps<-10
reps<-rep_sample_n(human_genes, size=100, replace=FALSE, reps=nsamps) %>% 
  group_by(replicate) %>% 
  summarise( # summaries
  MeanLength = mean(size),
  MedianLength = median(size),
  SDLength = sd(size),
  IQR = IQR(size)
) %>%
  ungroup

```
## Exercise: Making a distribution of sample means (from lab 7)

Step 4: Plot the sampling distribution with ggplot

```{r}
library(ggplot2)
#rerun previous code with nsamps=10000
ggplot(reps, aes(x=MeanLength)) + 
  geom_density() + 
  theme_minimal()

```

## Exercise: Making a distribution of sample means (from lab 7)

Final step: Obtain the grand mean (mean of sample means) and the standard deviation of the sample means and: compare with the population mean; calculate SEM using the distribution and the formula and compare.

```{r}
#Try it!

```


### Exercises!

Practice!

* Binomial hypothesis testing: `lab9_ex1.Rmd`
* Loops & Condtional Statements: Exercises in `lab9_ex2.Rmd`.

