---
title: "Hospital readmission rates of acute ischemic stroke in California"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

Tip: you can click on `visual` above this document to experience a friendlier interface.

## Motivation

According to the Nationwide Readmissions Database of the Healthcare Cost and Utilization Project between 2010 and 2015, the 30-day hospital readmission rate for acute ischemic stroke patients on a national level is 12.4% (Bambhroliya et al. 2018). A researcher wants to test whether the proportion of 30-day hospital readmissions for a California hospital with an "as expected" hospital quality rating differs from the national 30-day readmission proportion.

## Data

CSV Data file: [*readmin.csv*](input_files/readmin.cvs)

The data file contains ischemic stroke 30-day hospital readmission incidence data for a random sample of patients in a California hospital with an "as expected" quality rating obtained from a set of hospital records for 2014-2015. The 30-day readmission data from a sample of 50 patients was recorded.

Read in the dataset (it's in the input_files folder) which we'll call *readmin*. Use `head` to check what it looks like:

```{r}
# read dataset
# use head()
```

The variable *ReadmissionStatus* is a binary variable which is equal to 1 if the patient was readmitted to the hospital within 30 days of discharge and equal to 0 if the patient was not readmitted to the hospital within 30 days of discharge.

## Questions of interest

Two questions of interest are:

1.  Is the proportion of stroke patients readmitted within 30-days of discharge from CA hospital different from the nationwide proportion?

2.  What is the 95% confidence interval for the proportion of acute ischemic stroke patients readmitted within 30 days of discharge?

## Instructions

-   Install the `binom()` package and load it:

```{r}
# Install
# load
```

## Exploring the Data

Let's begin by looking at the variable of interest, readmission. Use the `table()` function.

```{r}
# Table of Readmission Status

```

*How many peope are readmitted to the hospital within 30 days? How many are not?*

## Data Analysis

Determine the proportion of individuals readmitted in the sample. *Tip: check out the `prop.table()` function.*

```{r}
# Proportion table of Readmission Status

```

Answer: of the sample was readmitted within 30 days.

#### The Binomial test

Write the null and alternative hypotheses for this statistical test, comparing the proportion of readmission in our study, to the population proportion of 0.124.

$H_0: \text{The proportion of readmission in the population is }______$

$H_A: \text{The proportion of readmission in the population is not }_______$

Run a two-sided binomial test to see if the sample proportion differs significantly from the population proportion. *Tip: Some possibilities were covered in lectures. You can also just check out the `binom` package now.*

```{r}
# Binomial test of readmission status

```

*What is your decision based on the binomial test?*

Answer:

#### Agresti-Coull 95% Confidence Interval for the proportion

Next, calculate the 95% CI for the proportion of stroke patients who were readmitted within 30 days of discharge using the Agresti-Coull method. *Tip: Some possibilities were covered in lectures. You can also just check out the `binom` package now.*

```{r}
# 95% CI for binomial test

```

*What is the Agresti-Coull 95% CI?*

Answer:

## Conclusions

Based on these results, we find no evidence that the population proportion is significantly different from 0.124. The 95% CI interval ($0.0525 < p < 0.2417$) indicates medium level of precision, given that it covers nearly 20% of possible proportions.

## References

Bambhroliya AB, Donnelly JP, Thomas EJ, et al., "Estimates and Temporal Trend for US Nationwide 30-Day Hospital Readmission Among Patients With Ischemic and Hemorrhagic Stroke.", *JAMA Netw Open*, 1[2018]:e181190 <https://jamanetwork.com/journals/jamanetworkopen/fullarticle/2696869>
