---
title: "Loopy for Loops, Crazy for Conditionals,  Silly for Samples - Hands-on!"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---
```{r setup, echo=F}
knitr::opts_chunk$set(echo=T, eval=T)
```
     
## Outline
Hello R students your silly student coder has returned to make your assignments notably worse than if Dr. B had made them. This week you will learn:

-intro to loops: for and while. (R also has a loop called repeat that I only learned about when I googled "types of loops in R. I don't think you'll need to know about it)
-intro to sampling
-as much dplyr as I can include to drive Levi crazy (ask Levi about his beef with tidyverse if you are unaware).

### What is a loop?
Loops are a programming element that repeat a portion of code a set number of times until the desired process is complete.

This guide someone wrote on the internet is really good if you want to also read it. https://intro2r.com/conditional-statements.html

#### For Loops
A for loop applies a command to each value provided then stops. basic example: 

```{r, echo=TRUE}
for (i in 1:5) {
  print(i)
}
```
You don't actually have to use i for the value there but most people do. 

So this loop printed each value for i. How would you write code that printed the values 6-10?

```{r for1}
for (i in 6:10) {
  print(i)
}
```

It's often useful to define a placeholder variable before running your loop. For example, this loop calculates the mean of a given data set. 

```{r}
numbers <- c(4, 22, 6, 13, 19, 2, 11)
sum <- 0
for(num in numbers) {
  sum <- sum + num
}
mean <- sum / length(numbers)
print(mean)

```


You can also write loops with character vectors. Here's a character vector. Write a loop that adds day to each day of the week.
days<-c("Mon", "Tues", "Wednes", "Thurs", "Fri", "Satur", "Sun")
(hint: if you add the argument sep = "" to paste you don't get a space in between)

```{r for2}
days<-c("Mon", "Tues", "Wednes", "Thurs", "Fri", "Satur", "Sun")

for (day in days){
 print(paste(day, "day", sep = ""))   
        
}


```


#### While Loops
A while loop repeats code until a condition is met. It's really important to include the condition because the loop might continue forever if you don't. 

Here's an example of a loop that prints x then multiplies it by two, as long as x is under 16. 
```{r}
x <- 2
while (x < 16) {
  print(x)
  x <- x * 2
}

```


Now write a while loop that prints the value, then doubles it, starting at 1 and  stopping at 100.

```{r while1}
x <- 1
while (x <= 100) {
  print(x)
  x <- x * 2
}


```

Just for fun try running that loop without with while(x). This will run forever. Make sure you remember to add the condition!


 Say you have a number and you want to find the smallest divisor that isn't 1. You could use a while loop to do this.  
```{r while2}

number <- 91
divisor <- 2
while (number %% divisor != 0) {
  divisor <- divisor + 1
}
print(divisor)


```
 





## Sampling
There is a basic function in R for taking samples very appropriately named sample. You need three arguments for it to work (there is a 4th but don't worry about it yet). The first is the data set, the second is the number of samples, and the third is with or without replacement. 

Read in the human_genes.csv file and name it human_genes. Play around with it using your strong exploratory data analysis skills. Get a good feel for the data set. 

Overall goal of this exercise is to compare sample means at different n to the population mean. 


Its big right? So let's take some samples. 
NOTE: Your output will look different to the pdf beacuse your samples will be different values. 

1. First filter it to only keep the size and name column. 
2. create a second data frame with summary statistics mean, median and SD of length. (what is interesting about this data??)
3. Now let's take some samples. Take 3 samples of gene size with replacement on. Use n = 10, 100, 1000. Calculate the mean of these samples, and save each mean as mean10, mean100, and mean1000. 
4. Compare these means to the population mean. Which is closest?


```{r samp1}
library(dplyr)
library(readr)
human_genes <- read_csv("fall_2022_materials/lab5/input_files/human_genes.csv") 


human_genes<-human_genes %>% select(name, size)

human_genes_summ <- human_genes %>%
  summarise(
    MeanLength = mean(size),
    MedianLength = median(size),
    SDLength = sd(size)
)

mean10<-mean(sample(human_genes$size, 10, replace = TRUE))
mean100<-mean(sample(human_genes$size, 100, replace = TRUE))
mean1000<-mean(sample(human_genes$size, 1000, replace = TRUE))
        
        
        
        
```



