---
title: "Loopy for Loops, Silly for Samples - Hands-on!"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output: pdf_document
#This code is from (jng, 2021) and is AMAZING! It automatically warps the text to fit the page
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---
```{r setup, echo=F}
knitr::opts_chunk$set(echo=T, eval=T)
```
     
## Outline
Hello R students your silly student coder has returned to make your assignments notably worse than if Dr. B had made them. This week you will learn:

-intro to loops: for (you will later learn about while and repeat)
-intro to sampling
-as much dplyr as I can include to drive Levi crazy (ask Levi about his beef with tidyverse if you are unaware).

### What is a loop?
Loops are a programming element that repeat a portion of code a set number of times until the desired process is complete. source (http://support.kodable.com/en/articles/417331-what-are-loops#:~:text=Definition%3A%20Loops%20are%20a%20programming,save%20time%20and%20minimize%20errors.)


This guide someone wrote on the internet is really good if you want to also read it. https://intro2r.com/conditional-statements.html


We are going to be focusing on for loops. A for loop applies a command to each value provided then stops. basic example: 

```{r, echo=TRUE}
for (i in 1:5) {
  print(i)
}
```
You don't actually have to use i for the value there but most people do. 

So this loop printed each value for i. How would you write code that printed the values 6-10?

```{r ex1}
for (i in 6:10) {
  print(i)
}
```

It's often useful to define a placeholder variable before running your loop. For example, this loop calculates the mean of a given data set. 

```{r}
numbers <- c(4, 22, 6, 13, 19, 2, 11)
sum <- 0
for(num in numbers) {
  sum <- sum + num
}
mean <- sum / length(numbers)
print(mean)

```


