---
title: "R Quiz 8"
output:
  ioslides_presentation:
    widescreen: true
    size: smaller
date: "2022-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE, eval=F)
data(iris)
```

## Fill in the code to return the output


```{r, eval=F}
survey_vector <- c("M", "F", "F")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")
___(factor_survey_vector)
```

```{r, eval=T}
survey_vector <- c("M", "F", "F")
factor_survey_vector <- factor(survey_vector)
levels(factor_survey_vector) <- c("Female", "Male")
summary(factor_survey_vector)
```

```{r}
View()
Summary()
summarise()
summary()
table()
```

## What is the correct way of removing column `Temp` from the `airquality` dataset?


```{r,eval=T, echo=F}
head(iris)

```

```{r}
iris %>% select(~Temp)
iris %>% select(rm(Temp))
iris %>% select(-Temp)
iris %>% select(!Temp)
```  
  


## Complete the code so that the conditional statement below prints "This number is divisible by 8" if it is divisible by 8 and "This number is not divisible by 8" if it isn't.

```{r, eval = F}

if (x ___ ) { #A
  ___ #B
} ___ { #C
  ___ #D
}
```  
  
Options: 
%% 8 == 0
 %% 8 == 0 
 %% 8 == 0
%% 8 == 0 
%% 8 ==0
 %% 8 ==0
%% 8 ==0 
 %% 8 ==0 
%% 8==0
 %% 8==0
 %% 8==0 
%% 8==0 
%%8== 0
 %%8== 0
 %%8== 0 
%%8== 0 
%%8==0
 %%8==0
%%8==0 
 %%8==0 
print("This number is divisible by 8")
print ("This number is divisible by 8")
else
 else
 else 
print("This number is not divisible by 8")
print ("This number is not divisible by 8")


## Examine the following code

```{r}
if (number < 10) {
  if (number < 5) {
    result <- "extra small"
  } else {
    result <- "small"
  }
} else if (number < 100) {
  result <- "medium"
} else {
  result <- "large"
}
print(result)

```

Which of the statements below are correct? 

If number is set to 6, “small” gets printed. T
If number is set to 100, "medium" is printed.
If number is set to 10 %% 5, R will generate an error because this is not a number. 
If number is set to 4, "extra small" gets printed. T
If number is set to 2500, R will generate an error, as result will not be defined.



## Predict the output of the following code:


```{r, eval=F}
my_function <- function(fname, lname) {
  paste(fname, lname)
}

my_function("Madonna")

```

```{r, echo=T, eval=F}
"Madonna" NULL
"Madonna"
An error because only one argument was provided, with no default for the other one
"Madonna" lname
An error because the argument name was not specified
```



## Predict the output of the following code

```{r}
Nested_function <- function(x, y) {
  if(!(x %% 2 == 0)){
  a <- x * y + 1
  } else {
    a <- x * y
  }
  return(a)
}

Nested_function(Nested_function(2,2), Nested_function(3,3))
```

45
An error
40
60