---
title: "Monty Hall sim"
author: "Levi"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: default
subtitle: Data types and data structures
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=T)
suppressMessages(library(dplyr))
suppressMessages(library(readr))
```


## Monty Hall

Summary

Three doors, one of which has a car and two have a goat

You pick one door and then the host reveals one of the two unpicked doors has a goat behind it and gives you an opportunity to swap

Example: you choose door 1 and the host reveals that door 3 has a goat. Do you swap to door two?

This is hard and unintuitive. Not swapping gives you a 1 in 3 chance of being correct, but does swapping raise your chances?

## Simulation as a way to solve unintuitive problems

Simulating the above scenario using code provides an easy way for us to test this hypothesis

For loops and sampling are particularly important and, using for loops, we will repeat the door picking 10,000 times. Unintuitive problems therefore can be solved by using code to simulate the situation over and over again.

This method has real world application: it is a useful way to test evolutionary hypotheses. Since we cannot directly observe evolution happening, we must simulate it! Likewise here, since its hard to directly understand why swapping results in 2/3 success to failure, simulation provides a test.

## Not swapping
```{r, eval=T}
#Defining the three doors as a new vector containing our choices, doors 1, 2, and 3.
doors <- 1:3
#an empty vector to store our results in
results <- rep(NA, 10000)
for(i in 1:10000){
  #initial choice of which door (1, 2, 3)
  choice <- sample(doors, 1)
  #which door is correct? 1, 2, or 3
  correct <- sample(doors, 1)
  #is our choice the sample as the correct door?
  results[i] <- choice==correct
}
```

## Results
Proportion of correct doors:
```{r, eval=T}
table(results)
print(paste("Proportion correct:", table(results)[2]/sum(table(results))))
```

## Swapping
```{r, eval=T}
for(i in 1:10000){#initial random choice of which door (1, 2, 3)
choice <- sample(doors, 1)
#which door is correct; random and independent from the choice 
correct <- sample(doors, 1)
#which door host reveals (which has to have a goat)
if(choice== correct){
  #from the doors, remove both the choice and correct positions
  wrongdoor <- sample(doors[-choice], size = 1)}
else{wrongdoor <- doors[-c(choice,correct)]}
#new choices not including the wrong door option
newchoices <- doors[-wrongdoor]
#here, we swap by removing our original choice from the new choices vector
finalchoice <- newchoices[-which(newchoices==choice)]
results[i] <- finalchoice==correct
}
```

## Swapping results
Proportion of correct doors:
```{r, eval=T}
table(results)
print(paste("Proportion correct:", table(results)[2]/sum(table(results))))
```
