---
title: "Lab 6"
author: "Bárbara Bitarello"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: default
subtitle: Data types and data structures
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=F)
suppressMessages(library(dplyr))
suppressMessages(library(readr))
human_genes<-read_csv("human_genes.csv")
```


## Today

* Quick recap

        - indexing elements (base R and dplyr)

* Go over some of lab 4

* New stuff!

        - Loops
        - Sampling


## Indexing elements: Base R

* Selecting a column (data.frame, tibble) or element (list) with `$`.
E.g. `iris` dataset

```{r, eval=T}
# select column containing petal lengths from iris
petal_lengths<-iris$Petal.Length
head(petal_lengths)
```

## Indexing elements: Base R

* Selecting a column (data.frame, tibble) or element (list) with `$`.
* E.g. a list containing `iris` and a vector containing all letters from the Roman alphabet

```{r, eval=T}
my_list<-list(iris=iris, letters=LETTERS)
str(my_list)
```

## Indexing elements: Base R

* Selecting a column (data.frame, tibble) or element (list) with `$`.
* E.g. a list containing `iris` and a vector containing all letters from the Roman alphabet

```{r, eval=T}
#index the second element of the list, i.e, the letters
#PS: $ only works if you name the elements
my_list$letters
```


## Indexing elements: Base R

Filtering elements with `[]`. E.g. data.frame

```{r, eval=T}
# Syntax: [row, col]
head(iris[iris$Sepal.Length>2,])

```

## Indexing elements: Base R

Filtering elements with `[]`. E.g. vector

```{r, eval=T}
# Syntax: [row, col]
LETTERS[LETTERS=='A']

LETTERS[3]

```
## Indexing elements: `dplyr`

Select columns with `select`:

```{r, eval=T}
#option 1
library(dplyr)
head(select(iris, Petal.Length))
```


## Indexing elements: `dplyr`

Select columns with `select`:

```{r, eval=T}
#option 2
iris %>% select(Petal.Length) %>% head()
```
## Indexing elements: `dplyr`

Filter rows with `filter`:

```{r, eval=T}
#option 1
head(filter(iris, Sepal.Length>2))
```

## Indexing elements: `dplyr`

Filter rows with `filter`:

```{r, eval=T}
#option 2
iris %>% filter(Petal.Length>2) %>% head()
```

## Loops in computer programming


* All programming languages have loops
* Loops are among the most basic and powerful of programming concepts. 
* A loop in a computer program is an instruction that repeats until a specified condition is reached. 
* The loop asks a question. If the answer requires action, it is executed. 
* The same question is asked again and again until no further action is required. 
* Each time the question is asked is called an iteration. 

## Loops in R

Three kinds of loops:

* `for` loop

* `while` loop 

* `repeat` loop [we won't discuss]

## `for` loops

Basic syntax

```{r, eval=F}
for (value in sequence){
  
  statement #DO SOMETHING
}

```

## Loops in R: `for` loops

```{r}
# iterate over elements of a sequence
for (i in 1: 4){
  
    print(i * 2)
  
}
```

## Loops in R: `for` loops

```{r}
#iterate over elements of a vector
for (i in c(-8^2, 2*4, 239, 29321)){

    print(i)
  
}
```

## Loops in R: `for` loops

```{r}
#create vector outside of loop and then use it to iterate
x <- c(-8^2, 2*4, 239, 29321)
for (i in x){
  
    print(i)
  
}
```

## `while` loops

Basic syntax

```{r, eval=F}
while ( condition ) {
  
  statement #do something
}

```


## Loops in R: `while` loops

```{r, eval=T}
y<-0
while (y <= 5){ #logical
    # statements
    print(y) # if y <=5, print it
    y = y + 1 #if y <=5 add 1 to it
}
y #print y after the loop has concluded
```

## `while` loops

A practical application: factorials. E.g. $5!$, which is $5\times 4\times 3\times 2\times 1$

```{r}
# whose factorial will be calculated
n <- 5 
factorial <- 1 # assigning the factorial variable
i <- 1 # and iteration variable to 1
# using while loop
while (i <= n){
    # multiplying the factorial variable
    # with the iteration variable
    factorial = factorial * i
    # incrementing the iteration variable
    i = i + 1
}
print(factorial) # displaying the factorial
```


## Sampling 

* In data analysis, taking samples of the data is the most common process done by the analysts. 
* To study and understand the data, sometimes taking a sample is the best way and it is mostly true in case of big data.
* R offers the standard function `sample()` to take a sample from the datasets. 
* Many business and data analysis problems will require taking samples from the data. 
* The random data is generated in this process with or without replacement.

## The `sample()` function

```{r,eval=F}
sample(x, size, replace = FALSE, prob = NULL)
```


* **x** - vector or a data set.
* **size** - sample size.
* **replace** - with or without replacement of values.
* **prob** - probability weights
