---
title: "Expected answers"
date: "`r Sys.Date()`"
output: pdf_document
#This code is from (jng, 2021) and is AMAZING! It automatically warps the text to fit the page
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval=T)
suppressMessages(library(tidyverse))
```

## Part 1: body mass and metabolic rate


**Ex.1)**

```{r ex1}
#install the tidyverse package here
#after you do this, you can comment the line so it doesn't reinstall this every time you knit your document.
#install.packages("tidyverse")
#then, load the tidyverse package here
library(tidyverse)
```

**Ex.2)**

Expected output

```{r ex2}
# type your code here
mass_metabolism<-read_csv("mass_metabolism.csv")
head(mass_metabolism)
```

Answer: It's a tibble. There are three columns.

**Ex.3)**

```{r ex3}
# type your code here
dim(mass_metabolism)
```

Answer: 25 rows, 3 columns

**Ex.4)**

```{r ex4}
# type your code here
is.tibble(mass_metabolism)
```


Answer: we had seen with `head` that this is a tibble. The `is.tibble` command confirms this by returning `TRUE`.

**Ex.5)**

```{r ex5}
# type your code here
str(mass_metabolism)
```

Answer:

* body_mass: numeric
* metabolic_rate: numeric
* family: character


**Ex.6)**

```{r ex6}
# type your code here
mass_metabolism2<-mass_metabolism %>% mutate(family=factor(family))
levels(mass_metabolism2$family)
mass_metab_levels<-levels(mass_metabolism2$family)
```

Answer: The levels for `mass_metabolism2$family` are:

"Antilocapridae" "Bovidae"        "Camelidae"      "Canidae"       
"Cervidae"       "Suidae"         "Tayassuidae"    "Tragulidae"    



**Ex.7)**

```{r ex7}
# type your code here
summary(mass_metabolism)
summary(mass_metabolism2)
```

Answer: `summary` is meaningless for a character column. Once we converted into factor, `summary` gives us counts within each category. For example, there are 11 rows where `family=='Bovidae'`.

**Ex.8)**

```{r ex8}
# type your code here
table(is.na(mass_metabolism2))
#the command above shows there are no missing data
mass_metab_na<-FALSE
```


Answer: there are no NAs.


**Ex.9)**


```{r ex9}
range_mass<-diff(range(mass_metabolism2$body_mass))
range_mass_bovidae<-mass_metabolism %>% filter(family=="Bovidae") %>% summarise(range=diff(range(body_mass)))
```

Answer: the range in mass overall is pretty huge (405387 kg!). But the range within Bovidae alone is also very impressive: 342800, so yes, I would say yes.

**Ex.10)**


```{r ex10}
mass_metabolism2 %>% filter(family=="Bovidae") %>% ggplot(aes(x=body_mass)) + geom_histogram()
```

Answer: skewed right.


**Ex.11)**


```{r ex11}
range_metab<-mass_metabolism %>% summarise(range=diff(range(metabolic_rate)))
range_metab_bovidae<-mass_metabolism %>% filter(family=="Bovidae") %>% summarise(range=diff(range(metabolic_rate)))
range_metab
range_metab_bovidae
```


Answer: Yes, the range in Bovidae alone (297) is almost as high as the overall range (302).


**Ex.12)**





```{r ex12}

mass_metabolism2 %>% filter(family=="Bovidae") %>% ggplot(aes(x=metabolic_rate)) + geom_histogram()
```



Answer: skewed right

**Ex.13)**



```{r ex13}
ggplot(mass_metabolism2, aes(x=body_mass, y=metabolic_rate)) + geom_point() + ylab("Metabolic rate") + xlab("Body mass")

```


Answer: body mass (explanatory), metabolic rate (response).

**Ex.14)**



```{r ex14}
ggplot(mass_metabolism2, aes(x=body_mass, y=metabolic_rate)) + geom_point(size=3) + ylab("Metabolic rate") + xlab("Body mass") + scale_y_continuous(trans='log10') + scale_x_continuous(trans='log10')


```




**Ex.15)**



```{r ex15}
ggplot(mass_metabolism2, aes(x=body_mass, y=metabolic_rate, color=family)) + geom_point(size=3) + ylab("Metabolic rate") + xlab("Body mass") + scale_y_continuous(trans='log10') + scale_x_continuous(trans='log10')
```



**Ex.16)**



```{r ex16}
#type your code here
ggplot(mass_metabolism2, aes(x=body_mass, y=metabolic_rate)) + geom_point(size=3) + ylab("Metabolic rate") + xlab("Body mass") + scale_y_continuous(trans='log10') + scale_x_continuous(trans='log10') + facet_wrap(~family, ncol=2)

ggplot(mass_metabolism2, aes(x=body_mass, y=metabolic_rate)) + geom_point(size=3) + ylab("Metabolic rate") + xlab("Body mass") + scale_y_continuous(trans='log10') + scale_x_continuous(trans='log10') + facet_wrap(~family, ncol=2)
```





**Ex.17)**


```{r ex17}
body_mass_summary<-mass_metabolism2 %>% group_by(family) %>% summarise(
  N=n(),
  Range_body_mass=diff(range(body_mass))
)
body_mass_summary
```

Answer: because there is only one value so range cannot really be calculated.

-----------

## Part 2: Carbon storage in shrubs


**Ex.18)**



```{r ex18}
# type your code here
shrubs<-read_csv("shrub-volume-data.csv")
head(shrubs)
dim(shrubs)
str(shrubs)
shrubs2<-shrubs %>% mutate(site=factor(site), experiment=factor(experiment))
summary(shrubs)
summary(shrubs2)
```
Answer: site and experiment were coded as `numeric` which does make sense since they are categorical variables. We change them to factors instead.



**Ex.19)**


```{r ex19}
#create shrubs3
shrubs3<-shrubs2 %>% na.omit()
#length
shrubs3 %>% ggplot(aes(x=length)) + geom_histogram(bins=10)
#width
shrubs3 %>% ggplot(aes(x=width)) + geom_histogram(bins=10)
# height
shrubs3 %>% ggplot(aes(x=height)) + geom_histogram(bins=10)
```



**Ex.20)**



```{r ex20}
shrubs3_summ<-shrubs3 %>% group_by(experiment) %>% summarise(N=n(), Range_length=diff(range(length)), Range_width=diff(range(width), Range_height=diff(range(height))))
shrubs3_summ
```

**Ex.21)**


```{r ex21}
#type your code here
shrubs4<-shrubs3 %>% mutate(area=width * length) %>% filter(height > 5) %>% arrange(length)
shrubs4
```



**Ex.22)**



```{r ex 22}
#type your code here
shrubs5<-shrubs3 %>% filter(height>4 & width>2)
shrubs5

```


**Ex.23)**





```{r ex23}
#type your code here

shrubs6<-shrubs3 %>% filter(experiment ==1 | experiment == 2)

shrubs6
```


**Ex.24)**



```{r ex24}
shrubs %>% filter(!is.na(height))
```

**Ex.25)**


```{r ex25}
shrub_volumes<-shrubs %>% mutate(area=width*length, volume=length*width*height)
shrub_volumes
```


