---
title: "Investigating the association between taking Ribavirin and death due to SARS in Singapore"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```
 
## Motivation

Ribavirin is a drug that has been used to treat various respiratory disease caused by viruses. It has a very few adverse side effects and works on a large variety of viruses. During the SARS outbreak, Ribavirin was used in several countries, including Singapore, as a treatment. Researchers were interested in measuring the effectiveness of Ribavirin on treating SARS.

## Data

CSV Data file: *sars.csv* (input files folder)

There were 229 individuals with SARS included in the retrospective study. Individuals either took Ribavirin or did not. The outcome of interest was whether an individual survived or died.

Read in the dataset, which we'll call *sars*.

```{r}
#install/load packages you will need
#read in the dataset
#do the basic checks
sars <- read.csv("sars.csv") 
head(sars) # Shows the first six rows of the data set
```

The variable *id* is an ID variable for each individual in the study. The variable *Ribavirin* is equal to *No Ribavirin* when an individual did not receive Ribavirin and equal to *Ribavirin* when an individual did received Ribavirin. The variable *Survival* is equal to *Died* when an individual died and equal to *Survived* when an individual survived.

## Questions of interest
Two questions of interest are:

1. Is there an association between the use of Ribavirin and survival?

2. What is the odds ratio of survival with Ribavirin use?


## Exploring the Data

Let's begin by looking at the two variables of interest, Ribavirin use and survival.

```{r}
# Table of Ribavirin use
table(sars$Ribavirin)

# Table of Survival
table(sars$Survival)

```

Of the 229 individuals included in the study:
a) How many received ribovarin? How many did not?
b) How many died? How many survived? 
c) Are there incorrect values or additional levels to either categorical variable? Is cleaning the data necessary?

## Data Analysis

### Investigating the association between Ribavirin use and survival

Begin by looking at a two-way table of Ribavirin use by survival.

```{r}


```

Describe what you see: how many individuals died/survived within each treatment group? 


This table does not give a complete view of the relationship because it does not show proportions.
Make a proportions table from the table you created above.

```{r}


```


You could also make a plot. A mosaic plot shows probabilities associated with survival and Ribavirin use. *Tip: there are at least two ways to do this - one with classic R plotting and one based on ggplot.*

```{r}

```
*Describe what your mosaic plot is showing you: does there seem to be an asymmetric distribution of individuals who dies amongst the two treatments?*

Answer: 

We can perform a $\chi^2$ contingency test to see if there is a statistically significant association between Ribavirin and survival. 

*Before you do this, how many degrees of freedom do you think you need here?*
Answer:


```{r}

```

*Describe the result of your hypothesis test: what is the p-value? Is it statistically significant? What do you conclude about the evidence of an association between Ribavirin use and death?*

Since the $P$-value is larger than 0.05, there is no statistically significant  evidence of an association between Ribavirin use and death.

*Are the results of the test above valid? Remember the assumptions for $\chi^2$ tests*?

* no more than 20\% of the cells can have an expected frequency less than five
* no cell can have an expected frequency less than one

```{r}

```

Answer:

### Odds ratio of survival

To better quantify this lack of association, we'll next calculate the odds ratio of survival with Ribavirin.

*What is the main difference between Fisher's exact test and the chi square contingency test we did above?*

Answer:

Calculate the OR of survival using the `fisher.test` function and obtain the OR of survival with Ribavirin and 95% CI for the OR:

```{r}
#Obtain the OR and 95% CI of the OR


```

*Describe your results: OR, CI of OR, what do they show and what do they mean.*
Answer: 

## Conclusions

What can you conclude about the effect of Ribavirin use on survival in patients with sars? Discuss any limitations you detected in this study.

## References
H-N. Leung, et al., "Investigational Use of Ribavirin in the Treatment of Severe Acute Respiratory Syndrome, Singapore, 2003", *Tropical Medicine and International Health*, 9 [2004]: 923-927
https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1365-3156.2004.01281.x

