---
title: "R Quiz 1"
output: 
  learnr::tutorial:
    progressive: TRUE
    allow_skip: FALSE
runtime: shiny_prerendered
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(knitr)
library(ggmosaic)
library(praise)
library(utils)
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
gradethis::gradethis_setup(
  pass.praise = TRUE,
  fail.encourage = TRUE,
  fail.hint = TRUE
)

```

## Vectors

### Exercise 1
1. Below, assign the number 9 to the variable `a`, and a vector with numbers, 2, 9, and 8, in that order, to the variable  `b`. Then, ask if `a` is equal to `b` and then multiply `a` and `b`.


```{r sim1, exercise=TRUE, allow_skip = FALSE}
               # assign 9 to a
```

```{r sim1-solution, allow_skip = FALSE}
a <- 9                
#Note: I am not sure how to allow multiple correct answers, say
# a<-9 and 
# a = 9
```


```{r sim1-check, allow_skip = FALSE}
grade_this_code()

```

### exercise 2
```{r sim2, exercise = TRUE, allow_skip = FALSE}
b <- c()            # assign a vector with 2, 9 and 8, in that order, to b
```


```{r sim2-solution, allow_skip = FALSE}
b <- c(2,9,8)      # assign a vector with 2, 9 and 8, in that order, to b 
```

```{r sim2-check, allow_skip = FALSE}
grade_this_code()
```


## Vector functions


### ex3 
```{r ex2setup}
a <- 9
b <- c(2,9,8)
```

```{r sim3, exercise = TRUE, exercise.setup = "ex2setup"}
                    # ask if a equals b 
```

```{r sim3-solution}
a == b               # ask if a equals b 
```

```{r sim3-check}
grade_this_code()
```


### exercise 4
```{r sim4, exercise = TRUE, exercise.setup = "ex2setup"}
                    # multiply a and b
```


```{r sim4-solution}
a * b             # multiply a and b 
```

```{r sim4-check}
grade_this_code()
```

## Quiz

```{r pRactice, echo=FALSE}
quiz(
    question("In the example above, what would R tell you when you ask it if a equals b?",
    answer("It doesn't return anything"),
    answer("`FALSE`"),
    answer("An error"),
    answer("The vector, `FALSE TRUE FALSE`", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement(), random_answer_order = TRUE
  ),
  question("In the example above, what does multiplying a and b return?",
    answer("The vector, `2 9 8`"),
    answer("The vector, `18 81 72`", correct = TRUE),
    answer("The vector, `18 9 8`"),
    answer("The vector, `4 81 64`"),
    answer("An error"),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement(), random_answer_order = TRUE
  ))
```

## R/Rstudio

```{r quiz2, echo=FALSE}
quiz(caption = "Quiz - Install R",
  question("Is R free to download and use?",
    answer("Yes!", correct = TRUE, message = "R is free _and_ open source, which means that anyone can read, use, and copy the source code of the R language."),
    answer("No.")
  )
  ,
    question("What is the RStudio IDE?",
    answer("An application that makes it easier to use R.", correct = TRUE, message = "RStudio organizes your workspace and makes it easier to write, use, debug, and save R code. I highly recommend it and will rely on the RStudio IDE in several tutorials."),
    answer("An application that let's you use R without writing any code", message = "And thank goodness! Code provides a reproducible record of your work, which is essential for data science."),
    answer("A spreadsheet program like Microsoft Excel."),
    answer("Another name for R", message = "R and RStudio are two separate things. R is a language, like English. Think of RStudio as a program that helps you use the language, kind of like how a word processing program helps you write in English."),
    allow_retry = TRUE
  ),
  question("Is the RStudio IDE free to download and use?",
    answer("Yes!", correct = TRUE, message = "Like R, RStudio is free and open-source. There do exist professional versions of RStudio with enhanced features, but we will not rely on those features in these tutorials."),
    answer("No.", message = "Like R, RStudio is free and open-source. There do exist professional versions of RStudio with enhanced features, but we will not rely on those features in these tutorials.")
  ))
```


## Use R to calculate:

a) The product of 15 by 17

```{r ex1, exercise=TRUE}

```

```{r ex1-check}
grade_result(
  pass_if(~identical(.result, 255))
)

```

b) Thirteen to the power of 3:

```{r ex2, exercise=TRUE}

```
```{r ex2-check}
grade_result(
  pass_if(~identical(.result, 2197))
)  
```  
c) Log of 14 (use the natural log):

```{r ex3, exercise=TRUE}

```

```{r ex3-check}
grade_result(
  pass_if(~identical(.result,  log(14)))
)  
```  

d) Log of 100 (use base 10):

```{r ex4, exercise=TRUE}

```

```{r ex4-check}
grade_result(
  pass_if(~identical(.result, 2))
)  
```  

e) The square root of 81:
```{r ex5, exercise=TRUE}

```

```{r ex5-check}
grade_result(
  #pass_if(~any(((9-0.5)<.result)==TRUE)&any(((9+0.5)>.result)==TRUE)
      
    pass_if(~identical(.result, 9))
)  
```  


Prove your reached the end and you're not a robot my creating a variable called `my_name` with your last name spelled backwards followed by your first name. E.g., mine would be "olleratib_arabrab".
 


```{r sandbox1, exercise=TRUE, eval = FALSE}
```


Last but not least, take a screenshot of this last task and email it to me with subject line: "R quiz 1"!

The end!


