---
title: "Types of Data Structures"
output: 
  learnr::tutorial:
    progressive: TRUE
    allow_skip: FALSE
runtime: shiny_prerendered
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(knitr)
library(ggmosaic)
library(praise)
library(utils)
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
gradethis::gradethis_setup(
  pass.praise = TRUE,
  fail.encourage = TRUE,
  fail.hint = TRUE
)

```

## Overview

Tasks: 

-   Work through this document
-   

Learning outcomes: 

1.  Learn about the different types of data

2.  Familiarize yourselves with common types of data structures

3.  Understand the interactions between vectors and matrices

4.  Use cbind and rbind

5.  Read about other ways of saving data (lists, dataframes, arrays)

## How many ways can data come

Numeric

This is the default computational data type in R. If you enter a number without a decimal point, R will automatically convert it into a numeric data type. Numeric values in R can hold any real number, including both integers and fractions, and even negative numbers. For example, 5, 23.5, and -3 are all numeric values.
```{r, echo=T, eval=F}
pi <- 3.14
print(pi)
```
Integer (for loops; we'll talk later)

This data type in R is used to hold whole numbers. Unlike numeric values, integers cannot have a decimal point. You can create an integer in R by adding an 'L' suffix to a number, like this: 5L. Without the 'L', R would treat the number as numeric, not integer.
```{r, echo=T, eval=F}
class_number <- 215L
```
Logical T/F
Logical data type in R is used to hold Boolean values: TRUE or FALSE. You can use logical operators like <, >, ==, etc. to generate logical values. For example, the operation 5 > 3 will return TRUE because 5 is indeed greater than 3.
```{r, echo=T, eval=F}
reality <- TRUE
```
Character

Character data type in R is used to hold strings of text. You can create a character string by enclosing text in quotes. For example, "Hello, World!" is a character string.
```{r, echo=T, eval=F}
types_of_data <- c("Numeric", "Interger","Logical", "Complex", "Character", "Raw") 
```

Complex <- 1 +2i

This data type in R is used to store complex numbers, which have both real and imaginary parts. The imaginary part is denoted by 'i'. For example, 2+3i is a complex number where 2 is the real part and 3 is the imaginary part. (We won't use this in the course)
```{r, echo=T, eval=F}
Complex <- 1 +2i #we dont know when you'd use this i guess if you are mathy idk
```

Raw (for the real nerds)

Raw data type in R is not commonly used for beginners. It is used to hold raw bytes, which can be anything from the contents of a file to an image. It's different from the other data types in that it doesn't represent something human-readable, but instead something that a machine might need to process. It's often used in more advanced topics like interfacing with other languages or dealing with binary data.
```{r, echo=T, eval=F}
hexcode1<- as.raw(0)
hexcode2 <- as.raw(128)
hexcode3 <- as.raw(255)
```

## Let's go through each type of data structure and why we use it

### Numeric 
Suppose you are studying the growth rate of a bacterial culture. You have the sizes of the bacterial colonies at different time points and you want to calculate the average growth rate.

```{r, echo=T, eval=F}
growth_data <- c(1.2, 2.3, 3.4, 5.6, 8.9) # Sizes of bacterial colonies at different time points

```


```{r setupnum}
growth_data <- c(1.2, 2.3, 3.4, 5.6, 8.9)
avg_growth<-mean(growth_data)
```
Suppose you want to take the average of these values. Paste avg_growth <- mean(growth_data) then print avg_growth
```{r examplenum, exercise = TRUE, exercise.setup = "setupnum"}
                    
```



```{r examplenum-solution}
avg_growth <- mean(growth_data) 
print(avg_growth)

```

```{r examplenum-check}
grade_this_code()
```

### Integer

You are counting the number of cells in different samples under a microscope. This is where integers come into play, as you can't have a fraction of a cell.

Suppose you are studying the growth rate of a bacterial culture. You have the sizes of the bacterial colonies at different time points and you want to calculate the average growth rate.

```{r, echo=T, eval=F}
cell_counts <- c(100, 200, 150, 175, 225) # Number of cells in different samples

```


```{r setupint}
cell_counts <- c(100, 200, 150, 175, 225)
total_cells <- sum(cell_counts)
```
Suppose you want to add up your total number of cells. You could define a variable total_cells <- sum(cell_counts). Paste the command and print the variable.

```{r exampleinteger, exercise = TRUE, exercise.setup = "setupint"}
 
                   
```


```{r xampleinteger-solution}
total_cells <- sum(cell_counts) 
print(total_cells)

```


```{r exampleinteger-check}
grade_this_code()
```


### Character

In bioinformatics, character strings are often used to represent DNA sequences.



```{r, echo=T, eval=F}
dna_sequences <- c("ATGC", "CGTA", "GATC", "TACG") # DNA sequences

```


```{r setupchr}
dna_sequences <- c("ATGC", "CGTA", "GATC", "TACG") # DNA sequences


```
Suppose you want to look at the first sequence from a long list. Paste the following code to define a character variable and then print it.
first_sequence <- dna_sequences[1] DNA sequence



```{r examplechr, exercise = TRUE, exercise.setup = "setupchr"}
                    
```


```{r examplechr-solution}
first_sequence <- dna_sequences[1] # Get the first DNA sequence
print(first_sequence)

```

```{r examplechr-check}
grade_this_code()
```


### Logical
```{r, echo=T, eval=F}
a <- c(1,2,3) #suppose this is your data and you misenter the first number
b <- c(2,2,3) #and you want to make sure your data matches previous result
which_entries_match <- a==b #paste this into the sandbox and call print() on the vector to look at the result
```


```{r setupbool}
a <- c(1,2,3) 
b <- c(2,2,3)

```


```{r examplebool, exercise = TRUE, exercise.setup = "setupbool"}
                    
```


```{r examplebool-solution}
which_entries_match <- a==b
print(which_entries_match)
```

```{r examplebool-check}
grade_this_code()
```


## Vectors and Matrices 

### Vectors
You already learned about vectors in the last lesson. They are a  sequence of data elements of the same basic type. They can be numeric, integer, complex, character, or logical. 

### Matrix

A matrix in R is a two-dimensional data structure, where each element is of the same atomic type. You can think of a matrix as a collection of vectors arranged side by side (for columns) or one on top of another (for rows). Here's an example of a numeric matrix. 

```{r matrix1, exercise = F }
numeric_matrix <- matrix(1:6, nrow = 2, ncol = 3) 
print(numeric_matrix)
```



## Interaction between Vectors and Matrices:

Vectors and matrices interact in various ways in R. Here are a few examples:

-You can create a matrix from vectors. Each vector can become a row or a column of the matrix.

-You can extract a row or a column from a matrix as a vector.

-You can use a vector to index a matrix, meaning to select certain rows or columns from the matrix.

-You can perform arithmetic operations between a matrix and a vector. In this case, R will often apply the operation element-wise, meaning it will operate on each corresponding pair of elements from the matrix and the vector. If the vector is shorter than the matrix, R will recycle the elements of the vector to match the length of the matrix.

### Creating a matrix from vectors:
You can use the cbind() function to combine vectors as columns of a matrix, or rbind() to combine vectors as rows.
```{r matrix2, exercise = F }
# Create two vectors
vector1 <- c(1, 2, 3)
vector2 <- c(4, 5, 6)

# Combine vectors as columns to create a matrix
matrix1 <- cbind(vector1, vector2)

# Combine vectors as rows to create a matrix
matrix2 <- rbind(vector1, vector2)

```
print them both to compare
```{r setupmat1}
vector1 <- c(1, 2, 3)
vector2 <- c(4, 5, 6)
matrix1 <- cbind(vector1, vector2)
matrix2 <- rbind(vector1, vector2)

```


```{r examplemat1, exercise = TRUE, exercise.setup = "setupmat1"}
                    
```



```{r examplemat1-solution}
print(matrix1)
print(matrix2)

```

```{r examplemat1-check}
grade_this_code()
```


### Extracting a row or a column from a matrix as a vector:
You can do this using indexing. In R, you use square brackets [] to index a matrix. The format is [row, column]

```{r mat2, exercise = F}
# Create a matrix
matrix3 <- matrix(1:9, nrow = 3)

# Extract the first row as a vector
row1 <- matrix3[1, ]

# Extract the first column as a vector
col1 <- matrix3[, 1]
```


```{r setupmat2}
matrix3 <- matrix(1:9, nrow = 3)
row1 <- matrix3[1, ]
col1 <- matrix3[, 1]
```

now print the matrix, row1 and col1 and examine

```{r examplemat2, exercise = TRUE, exercise.setup = "setupmat2"}
 


```



```{r examplemat2-solution}
 print(matrix3)    
  print(row1)
  print(col1)
```


```{r examplemat2-check}
grade_this_code()
```



### Performing arithmetic operations between a matrix and a vector:
R will apply the operation element-wise, and will recycle elements of the vector if necessary.


```{r mat3, exercise = F}
# Create a matrix
matrix4 <- matrix(1:6, nrow = 2)

# Create a vector
vector4 <- c(1, 2, 3)

# Add the vector to each row of the matrix
result4 <- matrix4 + vector4
```


```{r setupmat3}
matrix4 <- matrix(1:6, nrow = 2)
vector4 <- c(1, 2, 3)
result4 <- matrix4 + vector4

```

print the result

```{r examplemat3, exercise = TRUE, exercise.setup = "setupmat3"}
                    
```


```{r examplemat3-solution}
print(result4)
```

```{r examplemat3-check}
grade_this_code()
```
        

## Other ways you might see data








