---
title: "Hospital readmission rates of acute ischemic stroke in California"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```
 
## Chapters covered
 This covers chapters 3 and 7.
 
## Motivation

According to the Nationwide Readmissions Database of the Healthcare Cost and Utilization Project between 2010 and 2015, the 30-day hospital readmission rate for acute ischemic stroke patients on a national level is 12.4% (Bambhroliya et al. 2018). A researcher wants to test whether the proportion of 30-day hospital readmissions for a California hospital with an “as expected” hospital quality rating differs from the national 30-day readmission proportion.



## Data 

CSV Data file: *readmin.csv*

The  data file contains ischemic stroke 30-day hospital readmission incidence data for a random sample of patients in a California hospital with an “as expected” quality rating obtained from a set of hospital records for 2014-2015. The 30-day readmission data from a sample of 50 patients was recorded.

Read in the dataset, which we'll call *readmin*.

```{r}
readmin <- read.csv("readmin.csv") 
head(readmin) # Shows the first six rows of the data set
```

The variable *ReadmissionStatus* is a binary variable which is equal to 1 if the patient was readmitted to the hospital within 30 days of discharge and equal to 0 if the patient was not readmitted to the hospital within 30 days of discharge.

## Questions of interest
Two questions of interest are:

1. Is the proportion of stroke patients readmitted within 30-days of discharge from CA hospital different from the nationwide proportion?

2. What is the 95\% confidence interval for the proportion of acute ischemic stroke patients readmitted within 30 days of discharge?

## Instructions
Make sure to install the following packages using the install.packages() function, then load the corresponding libraries. 

```{r}
library(binom) 
```


## Exploring the Data
Let's begin by looking at the variable of interest, readmission.

```{r}
# Table of Readmission Status
table(readmin$ReadmissionStatus)
```

There are 6 people who are readmitted to the hospital within 30 days and 44 people who are not readmitted. 

## Data Analysis
Determine the proportion of individuals readmitted in the sample.
```{r}
# Proportion table of Readmission Status
prop.table(table(readmin$ReadmissionStatus))
```

12\% (or $p = \frac{6}{50} = 0.12$) of the sample was readmitted within 30 days. 

#### The Binomial test

Write the null and alternative hypotheses for this statistical test, comparing the proportion of readmission in our study, to the population proportion of 0.124.

$H_0: \text{The proportion of readmission in the population is }0.124$

$H_A: \text{The proportion of readmission in the population is not } 0.124$

Run a two-sided binomial test to see if the sample proportion differs significantly from the population proportion.

```{r}
# Binomial test of readmission status
binom.test(x = 6, n = 50, p = 0.124,
          alternative = "two.sided")
```

We fail to reject the null hypothesis that the population proportion differs significantly from 0.124 ($P$ = 0.9316).


#### Agresti-Coull 95\% Confidence Interval for the proportion

Next, calculate the 95\% CI for the proportion of stroke patients  who were readmitted within 30 days of discharge using the Agresti-Coull method.

```{r}
# 95% CI for binomial test
binom.confint(6, n = 50, method = "ac")
```

The true proportion of patients readmitted within 30 days is between 0.0525 and 0.2417 95% of the time that it is calculated from new data.

## Conclusions
Based on these results, we find no evidence that the population proportion is significantly different from 0.124. The 95% CI interval ($0.0525 < p < 0.2417$) indicates medium level of precision, given that it covers nearly 20% of possible proportions.  

## References
Bambhroliya AB, Donnelly JP, Thomas EJ, et al., "Estimates and Temporal Trend for US Nationwide 30-Day Hospital Readmission Among Patients With Ischemic and Hemorrhagic Stroke.", *JAMA Netw Open*, 1[2018]:e181190
https://jamanetwork.com/journals/jamanetworkopen/fullarticle/2696869